---
title: Diamond Deals
date: 2015-11-11
---

Looking to buy a diamond? This how-to guide describes three
steps I took to identify great deals on [Blue
Nile][blue-nile]. "Founded in 1999, Blue Nile has grown to become the
largest online retailer of certified diamonds and fine jewelry." The
code used in this analysis is available on [GitHub][repo]. This guide
proceeds as follows:

1.  download data from Blue Nile,
2.  model price as a function of diamond characteristics, and
3.  identify diamonds with extra low prices.

# Downloading Data

I've written a Python script to make downloading data from Blue Nile
easy. The script has been posted [here][download.py]. To download data
on all round diamonds on Blue Nile use the following command:

    python download.py --shape RD > my-diamonds.csv

For more information on the optional arguments the script accepts use:

    python download.py --help

Most of the download script is pretty easy to follow. Blue Nile is
using Apache Solr to serve JSON documents describing diamonds on
the site. The trickiest part is you can only get information on the
first 1,000 diamonds for each query; Blue Nile has limited how far we
can page through results. To work around this constraint, the download
script pages through results based on price. I only mention this if
you want to dig deeper into the download script.

```{r setup, message=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(msm)
library(knitr)
library(grDevices)

opts_chunk$set(echo=FALSE)
options(stringsAsFactors=TRUE)
diamonds <- read.csv('rounds.csv')
diamonds$tenths <- floor((10 * diamonds$carat) %% 10)

# Set up dummy variables
values <- list(
  cut=c('Good', 'Very Good', 'Ideal', 'Signature Ideal'),
  color=c('J', 'I', 'H', 'G', 'F', 'E', 'D'),
  clarity=c('SI2', 'SI1', 'VS2', 'VS1', 'VVS2', 'VVS1', 'IF', 'FL')
)
for (x in names(values)) {
  vals <- values[[x]]
  for (i in 2:length(vals)) {
    diamonds[[paste0(x, i)]] <- as.numeric(diamonds[[x]] == vals[i])
    diamonds[[x]] <- ordered(diamonds[[x]], levels=vals)
  }
}
dummies <- grep('\\d', names(diamonds), value=TRUE)

stats <- lapply(diamonds[, c('price', 'carat', dummies)], function(x)
  c(min=min(x), median=median(x), mean=mean(x), max=max(x))
)

my_format <- function(x) format(x, scientific=FALSE, big.mark=',')
```

On November 6, 2015, I downloaded data on all
`r my_format(nrow(diamonds))` round diamonds on Blue Nile. Below is a
plot of diamond price versus carat weight (both on log scales).

```{r big}
# Use some uniform noise to smooth out distribution of diamond weights.
# Diamond weight is rounded to 1/100 of a carat.
diamonds$noise <- runif(nrow(diamonds), 0, 0.01)

carats <- 2^(-2:4)
prices <- 1000 * 5^(0:4)
price_labels <- paste0('$', prices / 1000, ',000')
(
    ggplot(diamonds %>% filter(carat <= 8), aes(x=log(carat + noise), y=log(price))) +
    stat_smooth(method='lm') +
    geom_point(size=0.3) +
    scale_x_continuous(breaks=log(carats), labels=carats) +
    scale_y_continuous(breaks=log(prices), labels=price_labels) +
    ggtitle('Diamond price versus diamond weight') +
    ylab("Price") +
    xlab('Carat Weight') +
    theme_bw() +
    theme(panel.grid.minor.x=element_blank(), panel.grid.minor.y=element_blank())
)
```

# Modeling Price

Blue Nile's [buying guide][buying-guide] describes how the four C's
(cut, color, clarity, and carat weight) are the most important
characteristics when buying a diamond. It seems reasonable to model
price as a function of those four characteristics. Having played
around with the data bit, a multiplicative model seems like a good
choice. I model price as a product of carat weight raised to the power
$\beta$ times multipliers for the cut, color, and clarity of the
diamond
$$
price_i \propto carat_i^\beta \cdot cut_i \cdot color_i \cdot clarity_i.
$$
Taking $\log$'s of both sides allows this model to be estimated using
a linear regression
$$
\log(price_i) = \alpha + \beta \log(carat_i) + \delta_{cut_i} +
\delta_{color_i} + \delta_{clarity_i} + \epsilon_i.
$$
Focusing on diamonds weighing between 1.00 and 1.99 carats, we can see
the relationship between $\log(price_i)$ and $\log(carat_i)$ is
remarkably linear, with diamond color shifting the intercept but not
the slope of the relationship.

```{r zoomed-in}
carats <- seq(1, 1.9, 0.1)
prices <- 1000 * 2^(2:5)
price_labels <- paste0('$', prices / 1000, ',000')
diamonds <- diamonds %>% filter(1 <= carat & carat < 2)
(
    ggplot(diamonds, aes(x=log(carat + noise), y=log(price), color=color)) +
    guides(color = guide_legend(reverse = TRUE)) +
    geom_point(size=0.4) +
    stat_smooth(method='lm') +
    scale_x_continuous(breaks=log(carats), labels=carats) +
    scale_y_continuous(breaks=log(prices), labels=price_labels) +
    ggtitle('Diamond price as a function of weight and color') +
    ylab("Price") +
    xlab('Carat Weight') +
    theme_bw() +
    theme(panel.grid.minor.x=element_blank(),
    panel.grid.minor.y=element_blank(), legend.position=c(0.9, 0.2))
)

fstring <- paste('log(price) ~ log(carat) +',
  paste(dummies, collapse='+')
)
fit <- lm(fstring, data=diamonds)
```

Below is a summary of the fitted linear model. Generally, I put very
little weight on R-squared values, but this model explains 91.5% of
the observed variance in log price!

```{r}
summary(fit)
```

Exponentiating the coefficients from the regression model gives
estimates of the price multipliers associated with different diamond
characteristics. These multipliers can help a shopper decide what type
of diamond to consider. The omitted categories (cut = Good, color = J,
and clarity = SI2) have implicit coefficients of 0 and price
multipliers of 1. Is a G-color diamond worth
`r round(exp(coef(fit)['color4']), 2)` times the price of a J-color
diamond with the same cut, clarity, and carat weight?

```{r coefplot}
l <- lapply(dummies, function(s) {
  beta <- coef(fit)[[s]]
  variance <- vcov(fit)[s, s]
  new_variance <- deltamethod(~ exp(x1), beta, variance)
  c(multiplier=exp(beta), variance=new_variance)
})
multipliers <- data.frame(do.call(rbind, l))
multipliers$variable <- unlist(sapply(names(values), function(x) paste(values[[x]][-1], '-', x)))
multipliers$y <- nrow(multipliers) - 1:nrow(multipliers)
diff <- qnorm(0.95, mean=0, sd=sqrt(multipliers$variance))
multipliers$low <- multipliers$multiplier - diff
multipliers$high <- multipliers$multiplier + diff
(
    ggplot(multipliers, aes(x=multiplier, y=y)) +
    geom_point() +
    scale_y_continuous(breaks=multipliers$y, labels=multipliers$variable) +
    theme_bw() +
    geom_errorbarh(aes(xmax=high, xmin=low), height=0) +
    ylab('') +
    xlab('Price Multiplier') +
    ggtitle('Price multipliers for cut, color, and clarity types')
)
```

# Identifying Deals

Having read Blue Nile's buying guide a few times, they've convinced me
to care about all four of the four C's. When purchasing a diamond, the
following cut, color, and clarity are my baseline:

* $cut_i \ge$ Ideal: Represents roughly the top 3% of diamond quality
  based on cut. Reflects nearly all light that enters the diamond. An
  exquisite and rare cut.

* $color_i \ge$ H: Near-colorless. Color difficult to detect unless
  compared side-by-side against diamonds of better grades. An
  excellent value.

* $clarity_i \ge$ VS1: Very Slightly Included: Imperfections are not
  typically visible to the unaided eye. Less expensive than the VVS1
  or VVS2 grades.

Below I plot the diamonds that meet my baseline. Fitting a linear
relationship between $\log(price_i)$ and $\log(carat_i)$, I highlight
the best 1% of deals, the diamonds where the difference between
expected and actual price is greatest
$$
\alpha + \beta \log(carat_i) - \log(p_i) = -\epsilon_i.
$$

```{r deals}
focus <- diamonds %>% filter(1 <= carat, carat < 2, cut >= 'Ideal', color >= 'H', clarity >= 'VS1')
fit <- lm(log(price) ~ log(carat), focus)
focus$residual <- resid(fit)
focus$Deal <- focus$residual <= quantile(focus$residual, 0.01)
(
    ggplot(focus, aes(x=log(carat + noise), y=log(price))) +
    geom_point(aes(color=Deal, shape=Deal), size=1) +
    stat_smooth(method='lm') +
    ggtitle('Diamonds with low prices') +
    ylab("Price") +
    xlab('Carat Weight') +
    theme_bw() +
    theme(panel.grid.minor.x=element_blank(), panel.grid.minor.y=element_blank()) +
    scale_x_continuous(breaks=log(carats), labels=carats) +
    scale_y_continuous(breaks=log(prices), labels=price_labels) +
    scale_colour_manual(values=c('#dddddd', 'red')) +
    theme(legend.position="none")
)
```

The table below describes the top 10 diamonds found using my
criteria.

```{r}
focus <- focus %>% arrange(residual)
my_table <- focus %>% select(residual, cut, color, clarity, carat, price) %>% head(10)
kable(my_table, digits=2)
```

Disclaimer: This is one way to identify deals. A more general solution
would allow shoppers to enter preference parameters similar to the
regression coefficients found above. Taking preference parameters
$\beta$, the best deals would maximize the shopper's utility
$$
u(X_i, p_i) = X_i \beta - p_i.
$$

[blue-nile]: http://www.bluenile.com/
[solr]: http://lucene.apache.org/solr/
[buying-guide]: http://www.bluenile.com/education/diamonds/
[delta-method]: http://www.ats.ucla.edu/stat/r/faq/deltamethod.htm
[repo]: https://github.com/amarder/diamonds
[download.py]: https://github.com/amarder/diamonds/blob/master/download.py
